Structure constants and short-time simulations of open collective quantum spin systems

### 2019-03-26 soc_squeezing/collective_paper/collective_spin.tex
changed theory section to consider the evolution of a general Heisenberg operator

In this section we provide the basic theory for our method to compute
expectation values of collective spin operators, deferring lengthy
derivations to the appendices.  We consider systems composed of $N$
distinct spin-1/2 particles.  Defining individual spin-1/2 operators
$s_{n=\x,\y,\z}\equiv\sigma_n/2$ and
$s_\pm\equiv s_\x\pm is_\y=\sigma_\pm$ with Pauli operators
$\sigma_n$, we denote an operator that acts with $s_n$ on the spin
indexed by $j$ and trivially (i.e.~with the identity $\1$) on all
other spins by $s_n^{(j)}$.  We then define the collective spin
operators $S_n\equiv\sum_{j=1}^Ns_n^{(j)}$ for
$n\in\set{\x,\y,\z,+,-}$.  Identifying the set $\set{\S_{\v m}}$ as a
basis for all collective spin operators, with
$\v m\equiv\p{m_+,m_\z,m_-}\in\N_0^3$ and
$\S_{\v m}\equiv S_+^{m_+} S_\z^{m_\z} S_-^{m_-}$, we can expand all
collective spin Hamiltonians in the form
\begin{align}
  H = \sum_{\v m} h_{\v m} \S_{\v m},
  \label{eq:H_general}
\end{align}
where Hermiticity requires that $h_{\v m}^*=h_{\v m^*}$ for
$\v m^*\equiv\p{m_-,m_\z,m_+}$.  The time evolution of a particular
Heisenberg operator $\S_{\v m}$ under a Hamiltonian of the form in
Eq.~\eqref{eq:H_general} is then given by
\begin{align}
  \f{d}{dt}\S_{\v m}
  &= i\sum_{\v n}h_{\v n}\sp{\S_{\v n},\S_{\v m}}_-
  + \sum_\J \gamma_\J \D\p{\J;\S_{\v m}} \\
  &= \sum_{\v n} T_{\v m\v n} \S_{\v n},
  \label{eq:EOM}
\end{align}
where $\sp{X,Y}_\pm\equiv XY\pm YX$; $\J$ is a set of jump operators
with a corresponding decoherence rate $\gamma_\J$;
\begin{align}
  \D\p{\J;\O}
  \equiv \sum_{J\in\J} \p{J^\dag \O J - \f12\sp{J^\dag J,\O}_+}
\end{align}
is the Heisenberg picture dissipator, or Lindblad superoperator, for
all jump operators $J\in\J$; and $T_{\v m\v n}$ is a matrix element of
the time derivative operator.  Decoherence via uncorrelated decay of
individual spins at a rate $\gamma_-$, for example, would be described
by the set of jump operators $\J_-=\set{s_-^{(j)}:j=1,2,\cdots,N}$
with $\gamma_{\J_-}=\gamma_-$.  The commutator in Eq.~\eqref{eq:EOM}
can be determined from the structure constants $f_{\v\ell\v m\v n}$ of
the general product
$\S_{\v\ell}\S_{\v m}=\sum_{\v n}f_{\v\ell\v m\v n}\S_{\v n}$ worked
out in Appendices \ref{sec:identities}--\ref{sec:prod_general}, while
the effects of decoherence from jump operators (i.e.~elements of $\J$)
of the form $\gamma^{(j)}=\sum_n\gamma_ns_n^{(j)}$ and
$\Gamma=\sum_n\Gamma_nS_n$ are worked out in Appendices
\ref{sec:sandwich_single}--\ref{sec:decoherence_collective}.  We
consider these calculations to be the main technical contributions of
our work, with potential applications beyond the short-time simulation
method presented here.  These ingredients are sufficient to compute
matrix elements $T_{\v m\v n}$ of the time derivative operator $d/dt$
in Eq.~\eqref{eq:EOM} in most cases of practical interest.

The time derivative operator $d/dt$ will generally couple spin
operators $\S_{\v m}$ to spin operators $\S_{\v n}$ with higher
``weight'', i.e.~with $\abs{\v n}>\abs{\v m}$, where
$\abs{\v\ell}\equiv\sum_\lambda\ell_\lambda$.  The growth of operator
weight signifies the growth of many-body correlations.  In practice,
keeping track of this growth will eventually require more
computational resources than are available, meaning we must somehow
truncate our equations of motion.  The simplest truncation strategy
would be to take
\begin{align}
  \f{d}{dt} \S_{\v m}
  \to \sum_{w\p{\v n}<W} T_{\v m\v n} \S_{\v n}
  \label{eq:weight_truncation}
\end{align}
for some weight measure $w$, e.g.~$w\p{\v n}=\abs{\v n}$, and cutoff
$W$.  The truncation in Eq.~\eqref{eq:weight_truncation} closes the
system of differential equations defined by Eq.~\eqref{eq:EOM}, and
allow us to solve it using standard numerical methods.  Initial
conditions for this system of differential equations, i.e.~expectation
values of collective spin operators with respect to states that are
simple to prepare experimentally, are provided in Appendix
\ref{sec:initial_conditions}.

@@

We can expand Heisenberg operators in a Taylor series about time $t=0$
to write
\begin{align}
  \S_{\v m}\p{t}
  % = \left. \exp\p{t \f{d}{d\tau}} \S_{\v m}\p{\tau} \right|_{\tau=0}
  &= \sum_{k\ge0} \f{t^k}{k!}
  \left. \f{d^k}{d\tau^k} \S_{\v m}\p{\tau} \right|_{\tau=0} \\
  &= \sum_{k\ge0} \f{t^k}{k!}
  \sum_{\v n} T^k_{\v m\v n} \S_{\v n}\p{0},
  \label{eq:time_series}
\end{align}
where the matrix elements $T^k_{\v m\v n}$ of the $k$-th time
derivative are
\begin{align}
  T^0_{\v m\v n} &\equiv \delta_{\v m\v n}, \\
  T^1_{\v m\v n} &\equiv T_{\v m\v n}, \\
  T^{k>1}_{\v m\v n}
  &\equiv \sum_{\v p_1,\v p_2,\cdots,\v p_{k-1}}
  T_{\v m\v p_1} T_{\v p_1\v p_2} T_{\v p_2\v p_3}
  \cdots T_{\v p_{k-1}\v n},
\end{align}
with $\delta_{\v m\v n}=1$ if $\v m=\v n$ and 0 otherwise.  For
sufficiently short times, we can truncate the series in
Eq.~\eqref{eq:time_series} by taking
\begin{align}
  \S_{\v m}\p{t}
  \to \sum_{k=0}^M \f{t^k}{k!}
  \sum_{\v n} T^k_{\v m\v n} \S_{\v n}\p{0}.
  \label{eq:order_truncation}
\end{align}
Unlike the truncation in Eq.~\eqref{eq:weight_truncation}, the nonzero
matrix elements $T^k_{\v m\v n}$ for $k=0,1,\cdots,M$ now tell us
which operators are relevant for computing the expectation value
$\bk{\S_{\v m}\p{t}}$ to a fixed order $M$.  Note that the relation
$\S_{\v m^*}=\S_{\v m}^\dag$, which by Hermitian conjugation of
Eq.~\eqref{eq:EOM} also implies that
$T_{\v m^*\v n^*}=T_{\v m\v n}^*$, can be exploited to cut both the
number of initial-time expectation values $\bk{\S_{\v m}\p{0}}$ and
the number of matrix elements $T_{\v m\v n}$ that we may need to
explicitly compute roughly in half.

In principle, the factorial suppression of terms at higher orders of
the Taylor expansion in Eq.~\eqref{eq:time_series} implies for any
given time $t$, there exists a truncation order $M_t$ for which the
truncation error in Eq.~\eqref{eq:order_truncation} is negligibly
small.  In practice, only a maximal truncation order $M_{\t{max}}$ is
accessible with limited computational resources, such that the
truncated Taylor expansion in Eq.~\eqref{eq:order_truncation} only
allows computation of the correlator $\bk{\S_{\v m}\p{t}}$ to a
maximal time $t_{M_{\t{max}}}$.  As we will see in the following
section, $M_{\t{max}}=35$ will suffice for the computation of
collective spin correlators up to times that are relevant e.g.~for
spin squeezing protocols.

@@

The main lesson from Figure \ref{fig:benchmarking} is that the
truncated Taylor expansion in Eq.~\eqref{eq:order_truncation} gives
essentially exact results right up until a sudden and drastic
departure that is simple to diagnose by inspection.  This drastic
departure occurs because the truncated Taylor expansion neglects
high-weight (i.e.~large-$\abs{\v n}$) operators $\S_{\v n}$ whose
contributions to a correlator of interest $\bk{\S_{\v m}}$ eventually
become non-negligible.  For sufficiently large times $t$, these
individual contributions may generally be large compared to the value
of $\bk{\S_{\v m}}$, in which case their truncation will yield
non-physical results.  This breakdown mechanism is also the reason for
prolonged agreement between the truncated Taylor expansion and
benchmarking methods when squeezing the initial state $\ket{-\Z}$
rather than $\ket{\X}$: for initial states $\ket{-\Z}$, all
initial-time correlators $\bk{\S_{\v n}\p{0}}$ vanish unless
$n_+=n_-=0$ (see Appendix \ref{sec:initial_conditions}), so there is a
substantially smaller number of terms with non-zero contribution to
the correlator $\bk{\S_{\v m}}$, and therefore a substantially smaller
number of non-zero terms that are neglected in the truncated Taylor
expansion.

>>>>

In this section we provide the basic theory for our method to compute
expectation values of collective spin operators, deferring lengthy
derivations to the appendices.  We consider systems composed of $N$
distinct spin-1/2 particles.  Defining individual spin-1/2 operators
$s_{n=\x,\y,\z}\equiv\sigma_n/2$ and
$s_\pm\equiv s_\x\pm is_\y=\sigma_\pm$ with Pauli operators
$\sigma_n$, we denote an operator that acts with $s_n$ on the spin
indexed by $j$ and trivially (i.e.~with the identity $\1$) on all
other spins by $s_n^{(j)}$.  We then define the collective spin
operators $S_n\equiv\sum_{j=1}^Ns_n^{(j)}$ for
$n\in\set{\x,\y,\z,+,-}$.  Identifying the set $\set{\S_{\v m}}$ as a
basis for all collective spin operators, with
$\v m\equiv\p{m_+,m_\z,m_-}\in\N_0^3$ and
$\S_{\v m}\equiv S_+^{m_+} S_\z^{m_\z} S_-^{m_-}$, we can expand all
collective spin operators in the form
\begin{align}
  \O = \sum_{\v m} \O_{\v m} \S_{\v m}.
\end{align}
The time evolution of a Heisenberg operator $\O\p{t}$ under a
Hamiltonian $H$ is then given by
\begin{align}
  \f{d}{dt} \O\p{t}
  = T \O\p{t}
  = \sum_{\v m, \v n} \S_{\v m} T_{\v m\v n} \O_{\v n}\p{t}
  \label{eq:EOM}
\end{align}
for a time derivative (super-)operator $T$ with matrix elements
$T_{\v m\v n}$, determined by
\begin{align}
  T \O \equiv i \sp{H, \O}_- + \sum_\J \gamma_\J \D\p{\J; \O},
  \label{eq:time_deriv}
\end{align}
where $\sp{X,Y}_\pm\equiv XY\pm YX$; $\J$ is a set of jump operators
with a corresponding decoherence rate $\gamma_\J$; and
\begin{align}
  \D\p{\J;\O}
  \equiv \sum_{J\in\J} \p{J^\dag \O J - \f12\sp{J^\dag J,\O}_+}
\end{align}
is the Heisenberg dissipator, or Lindblad superoperator, for all jump
operators $J\in\J$.  Decoherence via uncorrelated decay of individual
spins at a rate $\gamma_-$, for example, would be described by the set
of jump operators $\J_-=\set{s_-^{(j)}:j=1,2,\cdots,N}$ with
$\gamma_{\J_-}=\gamma_-$.  The commutator in Eq.~\eqref{eq:time_deriv}
can be determined from the structure constants $f_{\v\ell\v m\v n}$ of
the general product
$\S_{\v\ell}\S_{\v m}=\sum_{\v n}f_{\v\ell\v m\v n}\S_{\v n}$ worked
out in Appendices \ref{sec:identities}--\ref{sec:prod_general}, while
the effects of decoherence from jump operators (i.e.~elements of $\J$)
of the form $\gamma^{(j)}=\sum_n\gamma_ns_n^{(j)}$ and
$\Gamma=\sum_n\Gamma_nS_n$ are worked out in Appendices
\ref{sec:sandwich_single}--\ref{sec:decoherence_collective}.  We
consider these calculations to be the main technical contributions of
our work, with potential applications beyond the short-time simulation
method presented here.  These ingredients are sufficient to compute
matrix elements $T_{\v m\v n}$ of the time derivative operator $T$ in
Eq.~\eqref{eq:time_deriv} in most cases of practical interest.

The time derivative operator $T$ will generally couple spin operators
$\S_{\v n}$ (i.e.~in $\O$) to spin operators $\S_{\v m}$ with higher
``weight'', i.e.~with $\abs{\v m}>\abs{\v n}$, where
$\abs{\v\ell}\equiv\sum_\alpha\ell_\alpha$.  The growth of operator
weight signifies the growth of many-body correlations.  In practice,
keeping track of this growth will eventually require more
computational resources than are available, meaning we must somehow
truncate our equations of motion.  The simplest truncation strategy
would be to take
\begin{align}
  \f{d}{dt} \O
  \to \sum_{w\p{\v m}<W} \S_{\v m}
  \sum_{\v n} T_{\v m\v n} \O_{\v n}
  \label{eq:weight_truncation}
\end{align}
for some weight measure $w$, e.g.~$w\p{\v m}=\abs{\v m}$, and cutoff
$W$.  The truncation in Eq.~\eqref{eq:weight_truncation} closes the
system of differential equations defined by Eq.~\eqref{eq:EOM}, and
allows us to solve it using standard numerical methods.  Initial
conditions for this system of differential equations, i.e.~expectation
values of collective spin operators with respect to states that are
simple to prepare experimentally, are provided in Appendix
\ref{sec:initial_conditions}.

@@

We can expand Heisenberg operators $\O\p{t}$ in a Taylor series about
the time $t=0$ to write
\begin{align}
  \O\p{t}
  = e^{t T} \O\p{0}
  % = \sum_{k\ge0} \f{t^k}{k!} T^k \O\p{0}
  = \sum_{k\ge0} \f{t^k}{k!}
  \sum_{\v m, \v n} \S_{\v m} T^k_{\v m\v n} \O_{\v n}\p{0},
  \label{eq:time_series}
\end{align}
where the matrix elements $T^k_{\v m\v n}$ of the $k$-th time
derivative are
\begin{align}
  T^0_{\v m\v n} &\equiv \delta_{\v m\v n}, \\
  T^1_{\v m\v n} &\equiv T_{\v m\v n}, \\
  T^{k>1}_{\v m\v n}
  &\equiv \sum_{\v p_1,\v p_2,\cdots,\v p_{k-1}}
  T_{\v m\v p_{k-1}} \cdots T_{\v p_3\v p_2}
  T_{\v p_2\v p_1} T_{\v p_1\v n},
\end{align}
with $\delta_{\v m\v n}=1$ if $\v m=\v n$ and 0 otherwise.  For
sufficiently short times, we can truncate the series in
Eq.~\eqref{eq:time_series} by taking
\begin{align}
  \O\p{t}
  \to \sum_{k=0}^M \f{t^k}{k!}
  \sum_{\v m, \v n} \S_{\v m} T^k_{\v m\v n} \O_{\v n}\p{0}.
  \label{eq:order_truncation}
\end{align}
Unlike the truncation in Eq.~\eqref{eq:weight_truncation}, the nonzero
matrix elements $T^k_{\v m\v n}$ for $k=0,1,\cdots,M$ now tell us
which operators are relevant for computing the expectation value
$\bk{\O\p{t}}$ to a fixed order $M$.  Note that the relation
$\S_{\v m}^\dag=\S_{\v m^*}$ for $\v m^*\equiv\p{m_-,m_\z,m_+}$, which
by Hermitian conjugation of Eq.~\eqref{eq:EOM} also implies that
$T_{\v m^*\v n^*}=T_{\v m\v n}^*$, can be exploited to cut both the
number of initial-time expectation values $\bk{\S_{\v m}}_{t=0}$ and
the number of matrix elements $T_{\v m\v n}$ that we may need to
explicitly compute roughly in half.

In principle, the factorial suppression of terms at higher orders of
the Taylor expansion in Eq.~\eqref{eq:time_series} implies for any
given time $t$, there exists a truncation order $M_t$ for which the
truncation error in Eq.~\eqref{eq:order_truncation} is negligibly
small.  In practice, only a maximal truncation order $M_{\t{max}}$ is
accessible with limited computational resources, such that the
truncated Taylor expansion in Eq.~\eqref{eq:order_truncation} only
allows computation of the expectation value $\bk{\O\p{t}}$ to a
maximal time $t_{M_{\t{max}}}$.  As we will see in the following
section, $M_{\t{max}}=35$ will suffice for the computation of
collective spin correlators up to times that are relevant e.g.~for
spin squeezing protocols.

@@

The main lesson from Figure \ref{fig:benchmarking} is that the
truncated Taylor expansion in Eq.~\eqref{eq:order_truncation} gives
essentially exact results right up until a sudden and drastic
departure that is simple to diagnose by inspection.  This drastic
departure occurs because the truncated Taylor expansion neglects
high-weight (i.e.~large-$\abs{\v m}$) operators $\S_{\v m}$ whose
contributions to a correlator of interest $\bk{\O\p{t}}$ eventually
become non-negligible.  For sufficiently large times $t$, these
individual contributions may generally be large compared to the value
of $\bk{\O\p{t}}$, in which case their truncation will yield
non-physical results.  This breakdown mechanism is also the reason for
prolonged agreement between the truncated Taylor expansion and
benchmarking methods when squeezing the initial state $\ket{-\Z}$
rather than $\ket{\X}$: for initial states $\ket{-\Z}$, all
initial-time correlators $\bk{\S_{\v m}}_{t=0}$ vanish unless
$m_+=m_-=0$ (see Appendix \ref{sec:initial_conditions}), so there is a
substantially smaller number of neglected terms with non-zero
contribution to the correlator $\bk{\O\p{t}}$.
